Fully Extrinsic Type Inference (DRAFT 2025-06-20)

### 2025-07-29 doc/extrinsic_type_inference_paper/paper.tex
update
update model typing rules for pattern matching; note negative occurence
update rules with negative position restriction

@@

    \inferrule {
      \exists e' .\ f\J{(}v\J{)} \rightsquigarrow e'
      \\
      v  \simeq p \given \sigma 
    } {
      (f\ \J{[}p\J{=>}e\J{]})\J{(}v\J{)} \rightsquigarrow e[\sigma]
    }

@@

@@

    \inferrule { 
      \delta \satisfies f \hastype \tau_l \J{->} \tau_r
    } {
      \delta \satisfies f\ \J{[}p \J{=>} e \J{]}\hastype \tau_l \J{->} \tau_r
    }

    \inferrule { 
      \forall \sigma \qua
      \delta \satisfies p[\sigma] \hastype \tau_l
      \implies
      \delta  \satisfies e[\sigma] \hastype \tau_r
      \\\\
      \forall \tau_l', \tau_r' \qua
      \delta \satisfies f \hastype \tau_l' \J{->} \tau_r'
      \implies
      \delta \satisfies \tau_l \subtypes \tau_l'
      \implies
      \delta \satisfies \tau_r' \subtypes \tau_r
    } {
      \delta \satisfies f\ \J{[}p \J{=>} e \J{]}\hastype \tau_l \J{->} \tau_r
    }

@@

    \inferrule { 
      \forall \delta' \qua
      \textbf{dom}(\delta') \subseteq \Theta \implies
      \delta \oplus \delta' \satisfies \Delta \implies 
      \delta \oplus \delta' \satisfies e \hastype \tau
      \\\\
      \exists \delta' \qua 
      \textbf{dom}(\delta') \subseteq \Theta \up 
      \delta \oplus \delta' \satisfies \Delta
    } {
      \delta \satisfies e \hastype \J{ALL[}\Theta\J{]}\Delta \J{:} \tau
    }

@@

@@

>>>>

\TODO{factor out definitions common to both dynamic and static systems into basic system}

\TODO{flip records, functions and other collections to associate right (matching Lean design)}

@@

    \inferrule {
      \nexists e' .\ f\J{(}v\J{)} \rightsquigarrow e'
      \\
      v  \simeq p \given \sigma 
    } {
      (f\ \J{[}p\J{=>}e\J{]})\J{(}v\J{)} \rightsquigarrow e[\sigma]
    }

@@

\begin{definition}
  \label{def:valuing}
  \emph{valuing}
  \hfill
  \small
  \boxed{\Downarrow e}
  \nopad
  \begin{mathpar}
    \inferrule { 
      \TODO{...}
    } {
      \Downarrow e 
    } 
  \end{mathpar}
\end{definition}

\TODO{create a seperate relation restricted negations of model typing}

\TODO{use antec rules to compensate for non-self reference in negative position}

\TODO{consider renaming to dynamic typing}

@@

@@

@@

    \inferrule { 
      \forall \delta' \qua
      \textbf{dom}(\delta') \subseteq \Theta \implies
      \delta \oplus \delta' \VDash \Delta \implies 
      \delta \oplus \delta' \satisfies e \hastype \tau
      \\\\
      \exists \delta' \qua 
      \textbf{dom}(\delta') \subseteq \Theta \up 
      \delta \oplus \delta' \satisfies \Delta
    } {
      \delta \satisfies e \hastype \J{ALL[}\Theta\J{]}\Delta \J{:} \tau
    }

    \inferrule { 
      \TODO{diff antec}
      \\\\
      p \Uparrow \eta 
      \\
      \delta \satisfies f \hastype \tau_l \J{->} \tau_r
    } {
      \delta \satisfies \J{[}p \J{=>} e \J{]}\ f \hastype \tau_l \typdiff \eta \J{->} \tau_r
    }

    \inferrule { 
      \TODO{needs separate inter antec}
      \\\\
      \forall \sigma \qua 
      \delta \VDash p[\sigma] \hastype \tau_l
      \implies
      \delta  \satisfies e[\sigma] \hastype \tau_r
    } {
      \delta \satisfies \J{[}p \J{=>} e \J{]}\ f\hastype \tau_l \J{->} \tau_r
    }

    \inferrule { 
      \TODO{inter antec left}
      \\\\
      \delta \satisfies e \hastype \tau_l \J{->} \tau
    } {
      \delta \satisfies e \hastype (\tau_l \J{\&} \tau_r) \J{->} \tau
    }

    \inferrule { 
      \TODO{inter antec right}
      \\\\
      \delta \satisfies e \hastype \tau_r \J{->} \tau
    } {
      \delta \satisfies e \hastype (\tau_l \J{\&} \tau_r) \J{->} \tau
    }

    \inferrule { 
      \TODO{all antec}
      \\\\
      \textbf{dom}(\delta') \subseteq \Theta 
      \\
      \delta \oplus \delta' \satisfies \Delta 
      \\
      \delta \oplus \delta' \satisfies e \hastype \tau_l \J{->} \tau_r
    } {
      \delta \satisfies e \hastype (\J{ALL[}\Theta \J{]}\Delta\ \tau_l) \J{->} \tau_r
    }

    \inferrule { 
      \TODO{union antec}
      \\\\
      \delta \satisfies e \hastype \tau_l \J{->} \tau
      \\
      \delta \satisfies e \hastype \tau_r \J{->} \tau
    } {
      \delta \satisfies e \hastype (\tau_l \J{|} \tau_r) \J{->} \tau
    }

    \inferrule { 
      \TODO{exi antec}
      \\\\
      \forall \delta' \qua 
      \textbf{dom}(\delta') \subseteq \Theta
      \implies
      \delta \oplus \delta' \VDash \Delta 
      \implies
      \delta \satisfies e \hastype \tau_l \J{->} \tau_r
    } {
      \delta \satisfies e \hastype (\J{EXI[} \Theta \J{]}\Delta\ \tau_l) \J{->} \tau_r
    }

    \inferrule { 
      \TODO{lfp antec}
      \\\\
      \alpha \notin \textbf{ftv}(\tau_l[\alpha\slash\J{@}]^+)
      \\
      \delta\ \alpha \slash (\J{LFP[}\alpha \J{]} \tau_l) \satisfies e \hastype \tau_l \J{->} \tau_r
    } {
      \delta \satisfies e \hastype (\J{LFP[}\alpha \J{]} \tau_l) \J{->} \tau_r
    }

@@

    % \inferrule { 
    %   \TODO{remove unflipped}
    %   \\\\
    %   \delta \satisfies f \hastype \tau_l \J{->} \tau_r
    % } {
    %   \delta \satisfies f\ \J{[}p \J{=>} e \J{]}\hastype \tau_l \J{->} \tau_r
    % }

    % \inferrule { 
    %   \TODO{remove unflipped}
    %   \\\\
    %   \forall \sigma \qua
    %   \delta \satisfies p[\sigma] \hastype \tau_l
    %   \implies
    %   \delta  \satisfies e[\sigma] \hastype \tau_r
    %   \\\\
    %   \forall \tau_l', \tau_r' \qua
    %   \delta \satisfies f \hastype \tau_l' \J{->} \tau_r'
    %   \implies
    %   \delta \satisfies \tau_l \subtypes \tau_l'
    %   \implies
    %   \delta \satisfies \tau_r' \subtypes \tau_r
    % } {
    %   \delta \satisfies f\ \J{[}p \J{=>} e \J{]}\hastype \tau_l \J{->} \tau_r
    % }

    % \inferrule { 
    %   \delta \satisfies f \hastype \tau_l \J{->} \tau_r
    %   \\\\
    %   \forall \tau_l', \tau_r' \qua
    %   \delta \satisfies \J{[}p \J{=>} e \J{]} \hastype \tau_l' \J{->} \tau_r'
    %   \implies
    %   \delta \satisfies \tau_l \subtypes \tau_l'
    %   \implies
    %   \delta \satisfies \tau_r' \subtypes \tau_r
    % } {
    %   \delta \satisfies \J{[}p \J{=>} e \J{]}\ f \hastype \tau_l \J{->} \tau_r
    % }

    % \inferrule { 
    %   \TODO{how to deal with recursive negative occurence?}
    %   \\\\
    %   \forall v \qua 
    %   \delta \satisfies v \hastype \tau_l
    %   \implies
    %   \exists \sigma \qua
    %   v \simeq p \given \sigma 
    %   \up
    %   \delta  \satisfies e[\sigma] \hastype \tau_r
    % } {
    %   \delta \satisfies \J{[}p \J{=>} e \J{]}\ f\hastype \tau_l \J{->} \tau_r
    % }

    % \inferrule { 
    %   \TODO{how to deal with recursive negative occurence?}
    %   \\\\
    %   \TODO{intersection elim is built into pattern matching}
    %   \\\\
    %   \forall v \qua 
    %   \delta \VDash v \hastype \rho
    %   \implies
    %   \exists \sigma \qua
    %   v \simeq p \given \sigma 
    %   \up
    %   \delta  \satisfies e[\sigma] \hastype \tau_r
    % } {
    %   \delta \satisfies \J{[}p \J{=>} e \J{]}\ f\hastype \rho \J{->} \tau_r
    % }
